---
output: 
  html_fragment:
    keep_md: true
    smart: true
    md_extensions: +autolink_bare_uris+tex_math_single_backslash
params:
    # lib_dir: /home/lionnetc/travail/document/DRYVER/bact02
    lib_dir: /bettik/lionnetc/dryver/bact02_merged/run_clean_report
    rds_path: rds_files/bact02_clean_clustered_8.rds
    clust_rds_path: rds_files/bact02_clean_motus_9.rds
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, echo=F}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, load-dependencies, echo=F, message=FALSE, warning=FALSE, results='hide'}
.libPaths( c( .libPaths(), "/bettik/lionnetc/my_r_libraries/") )
library("metabaR")
library("ggplot2")
library("cowplot")
library("reshape2")
library("ade4")
library("vegan")
library("ape")
library("ggpubr")
library("stringr")
library("gridExtra")
```


```{r init-variable, message=FALSE, warning=FALSE}
lib_dir <- params$lib_dir
rds_file <- file.path(lib_dir, params$rds_path)
clust_rds_file <- file.path(lib_dir, params$clust_rds_path)
```

```{r, function, echo=F, warning=F, message=F}
## Function
ggBoxSamplesMtd <- function(metabarlist, sample_mtd){
  check <- melt(data.frame(metadata=metabarlist$samples[,sample_mtd],
                           nb_reads=as.vector(rowSums(metabarlist$reads)), 
                           nb_motus=as.vector(rowSums(metabarlist$reads>0)), 
                           stringsAsFactors = F), id.vars="metadata")
  
  ggplot(check, aes(x=metadata, y=value, color=metadata)) + 
    geom_boxplot() + theme_bw() + 
    geom_jitter(alpha=0.2) + 
    facet_wrap(~variable, scales = "free_y") + 
    theme(axis.text.x = element_text(angle=45, h=1)) +
    ggtitle("Distribution of reads and motus accross samples metadata")
}

ggBoxPCRMtd <- function(metabarlist, sample_mtd){
  check <- melt(data.frame(metadata=metabarlist$pcrs[,sample_mtd],
                           nb_reads=as.vector(rowSums(metabarlist$reads)), 
                           nb_motus=as.vector(rowSums(metabarlist$reads>0)), 
                           stringsAsFactors = F), id.vars="metadata")
  
  ggplot(check, aes(x=metadata, y=value, color=metadata)) + 
    geom_boxplot() + theme_bw() + 
    geom_jitter(alpha=0.2) + 
    facet_wrap(~variable, scales = "free_y") + 
    theme(axis.text.x = element_text(angle=45, h=1)) +
    guides(color=guide_legend(ncol=2)) +
    ggtitle("Distribution of reads and motus accross pcrs metadata")
}

## this function keep only the rank containning in the rank_vector and format the taxonomic lineage
filter_on_rank <- function(taxo_path, rank_vector){
  rank_filter <- lapply(taxo_path, str_detect, rank_vector);
  filtered_matrix <- matrix(unlist(rank_filter), 
                            nrow=length(unlist(rank_filter[1])))
  filtered_res <- as.vector(taxo_path[apply(filtered_matrix, 2, any)])
  available_rank <- apply(filtered_matrix, 1, any)
  new_path <- lapply(seq_along(available_rank),
                     function(ind){
                       ranknames <- gsub(".*@", "@", filtered_res)
                       if (rank_vector[ind] %in% ranknames) {
                         filtered_res[which(rank_vector[ind]==ranknames)]}
                       else {
                         indn <- 1
                         for(index in ind:1){
                           if(rank_vector[index] %in% ranknames){
                             indn <- which(rank_vector[index]==ranknames)
                             break
                           }
                         }
                         # paste the last taxon name with the active rank and undefined
                         paste(gsub("@.*", "", filtered_res[indn]), 
                               paste("(", 
                                     paste(gsub("@", "", rank_vector[ind]),
                                           "undefined", 
                                           sep=" "),
                                     ")", 
                                     sep=""), 
                               sep=" ")
                       }
                     })
  new_path
}

## format all ecotag path
format_ecotag_path <- function(path, ranks=rank){
  rank_vector <- paste("@", ranks, sep="")
  tmp_data <- strsplit(path, ":")
  filtered_data <- lapply(tmp_data, filter_on_rank, rank_vector)
  
  formated_data <- lapply(filtered_data, 
                          function(taxo_value){
                            gsub("@.*", "", as.character(taxo_value))})
  out <- data.frame(t(matrix(unlist(formated_data), 
                                nrow=length(unlist(formated_data[1])))), stringsAsFactors = F)
  names(out) <- ranks
  out
}
```


```{r load-initial-data, warning=FALSE, message=F, echo=F}
mtbl <- readRDS(rds_file)

sample_rep <- subset_metabarlist(mtbl, "pcrs", mtbl$pcrs$type=="sample")
samples <- aggregate_pcrs(sample_rep, FUN = FUN_agg_pcrs_sum)

rm(sample_rep)
rm(mtbl)
gc_stat <- gc()
```

Abundance is normalized with the Hellinger method : (sqrt(x/nb_total_in_samples))
```{r, prepare-sample-for-comparative-analysis, warning=F}
samples_freq <- samples
samples_freq$reads <- decostand(samples_freq$reads, "hellinger")
```

## Motus - unique sequence {.tabset}
### Boxplot {.tabset}
#### DNA plate
```{r, boxplot-by-mtd-dnaplate, fig.height=8, fig.width=10, warning=F}
ggBoxPCRMtd(samples, "dna_plate")
```

#### Sample type
```{r, boxplot-by-mtd-soil, fig.height=8, fig.width=8, warning=F}
ggBoxSamplesMtd(samples, "Sample_type")
```

#### Country 
```{r, boxplot-by-mtd-Country, fig.height=8, fig.width=8, warning=F}
ggBoxSamplesMtd(samples, "Country")
```

#### Flow regimen 
```{r, boxplot-by-mtd-flow, fig.height=8, fig.width=8, warning=F}
ggBoxSamplesMtd(samples, "Flow_regimen")
```


### PCA - Abundance {.tabset}
```{r, pca-by-sample-all, fig.height=10, fig.width=14, warning=F, results='asis', warning=F, message=F}
res.pca <- dudi.pca(samples_freq$reads, scannf = F, nf=4, scale = F)
categories <- samples_freq$samples[rownames(res.pca$li), "Sample_type"]
categories2 <- as.character(samples_freq$samples[rownames(res.pca$li), "Country"])
categories3 <- as.character(samples_freq$samples[rownames(res.pca$li), "Flow_regimen"])
```

#### All - Country
```{r, pca-by-sample-all-Country, fig.height=12, fig.width=14, warning=F, warning=F, message=F}
g1 <- ggplot(res.pca$li,
       aes(x = Axis1, y = Axis2, color=categories2, shape=categories)) +
  geom_point(size = 2.5) + 
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PCA by sample - Bact02") + 
  guides(color=guide_legend(ncol=2))

g2 <- ggplot(res.pca$li,
       aes(x = Axis1, y = Axis3, color=categories2, shape=categories)) +
  geom_point(size = 2.5) + 
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PCA by sample - Bact02") + 
  guides(color=guide_legend(ncol=2))

grid.arrange(g1, g2)

rm(g1)
rm(g2)
```

#### All - Sample type
```{r, pca-by-sample-all-soil, fig.height=12, fig.width=14, warning=F, message=F}
g1 <- ggplot(res.pca$li,
       aes(x = Axis1, y = Axis2, color=categories, shape=categories2)) +
  geom_point(size = 2.5) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PCA by sample - Bact02") +
  guides(color=guide_legend(ncol=2))


g2 <- ggplot(res.pca$li,
       aes(x = Axis1, y = Axis3, color=categories, shape=categories2)) +
  geom_point(size = 2.5) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PCA by sample - Bact02") +
  guides(color=guide_legend(ncol=2))

grid.arrange(g1, g2)

rm(g1)
rm(g2)
```

#### All - Flow regimen
```{r, pca-by-sample-all-flow-regimen, fig.height=12, fig.width=14, warning=F, message=F}
g1 <- ggplot(res.pca$li,
       aes(x = Axis1, y = Axis2, color=categories, shape=categories3)) +
  geom_point(size = 2.5) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PCA by sample - Bact02") +
  guides(color=guide_legend(ncol=2))


g2 <- ggplot(res.pca$li,
       aes(x = Axis1, y = Axis3, color=categories, shape=categories2)) +
  geom_point(size = 2.5) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PCA by sample - Bact02") +
  guides(color=guide_legend(ncol=2))

grid.arrange(g1, g2)

rm(res.pca)
rm(categories)
rm(categories2)
rm(categories3)
rm(g1)
rm(g2)
```


#### Country {.tabset}
```{r, pca-by-sample-Country, fig.height=12, fig.width=14, results='asis', warning=F, message=F}
for(country in sort(unique(samples_freq$samples$Country))){
  cat(paste0("##### ", country, "\n"))
  
  sub_mat <- subset_metabarlist(samples_freq, "samples", samples_freq$samples$Country==country)
  res.pca <- dudi.pca(sub_mat$reads, scannf = F, nf=4, scale = F)
  categories <- sub_mat$samples[rownames(res.pca$li), "Flow_regimen"]
  categories2 <- as.character(sub_mat$samples[rownames(res.pca$li), "Sample_type"])
  
  g1 <- ggplot(res.pca$li,
         aes(x = Axis1, y = Axis2, color=categories2, shape=categories)) +
    geom_point(size = 2.5) + 
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle(paste0("PCA by sample - Bact02 - ", country)) + 
    guides(color=guide_legend(ncol=2))
  
  g2 <- ggplot(res.pca$li,
         aes(x = Axis1, y = Axis3, color=categories2, shape=categories)) +
    geom_point(size = 2.5) + 
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle(paste0("PCA by sample - Bact02 - ", country)) + 
    guides(color=guide_legend(ncol=2))
  
  grid.arrange(g1, g2)
  cat("\n\n")
  
  rm(res.pca)
  rm(categories)
  rm(categories2)
  rm(g1)
  rm(g2)
  rm(sub_mat)
}
```


#### Sample type {.tabset}
```{r, pca-by-sample-sample-type, fig.height=12, fig.width=14, warning=F, results='asis', message=F}
for(sample_type in sort(unique(samples_freq$samples$Sample_type))){
  sub_mat <- subset_metabarlist(samples_freq, "samples", samples_freq$samples$Sample_type==sample_type)
  cat(paste0("##### Sample type - ", sample_type, "\n"))
  res.pca <- dudi.pca(sub_mat$reads, scannf = F, nf=4, scale = F)
  categories <- sub_mat$samples[rownames(res.pca$li), "Flow_regimen"]
  categories2 <- sub_mat$samples[rownames(res.pca$li), "Country"]
  
  g1 <- ggplot(res.pca$li,
         aes(x = Axis1, y = Axis2, color=categories2, shape=categories)) +
    geom_point(size = 2.5) + 
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle(paste0("PCA by sample - Bact02 - Sample type: ", sample_type)) + 
    guides(color=guide_legend(ncol=2))
  
  
  g2 <- ggplot(res.pca$li,
         aes(x = Axis1, y = Axis3, color=categories2, shape=categories)) +
    geom_point(size = 2.5) + 
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle(paste0("PCA by sample - Bact02 - Sample type: ", sample_type)) + 
    guides(color=guide_legend(ncol=2))
  
  grid.arrange(g1, g2)
  cat("\n\n")
  
  rm(res.pca)
  rm(categories)
  rm(categories2)
  rm(g1)
  rm(g2)
  rm(sub_mat)
}
```

#### Flow regimen {.tabset}
```{r, pca-by-sample-flow, fig.height=12, fig.width=14, warning=F, results='asis', message=F}
for(flow in sort(unique(samples_freq$samples$Flow_regimen))){
  sub_mat <- subset_metabarlist(samples_freq, "samples", samples_freq$samples$Flow_regimen==flow)
  cat(paste0("##### Flow regimen - ", flow, "\n"))
  res.pca <- dudi.pca(sub_mat$reads, scannf = F, nf=4, scale = F)
  categories <- sub_mat$samples[rownames(res.pca$li), "Sample_type"]
  categories2 <- sub_mat$samples[rownames(res.pca$li), "Country"]
  
  g1 <- ggplot(res.pca$li,
         aes(x = Axis1, y = Axis2, color=categories2, shape=categories)) +
    geom_point(size = 2.5) + 
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle(paste0("PCA by sample - Bact02 - Flow regimen: ", flow)) + 
    guides(color=guide_legend(ncol=2))
  
  
  g2 <- ggplot(res.pca$li,
         aes(x = Axis1, y = Axis3, color=categories2, shape=categories)) +
    geom_point(size = 2.5) + 
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle(paste0("PCA by sample - Bact02 - Flow regimen: ", flow)) + 
    guides(color=guide_legend(ncol=2))
  
  grid.arrange(g1, g2)
  cat("\n\n")
  
  rm(res.pca)
  rm(categories)
  rm(categories2)
  rm(g1)
  rm(g2)
  rm(sub_mat)
}
```


### PCoA - Presence-Absence {.tabset}
```{r, pcoa-by-sample-all, fig.height=10, fig.width=14, warning=F, results='asis', message=F}
samples_PA <- samples$reads
samples_PA[samples_PA>0] <- 1
samples_PA <- samples_PA[, colSums(samples_PA)>1]
samples_PA <- samples_PA[rowSums(samples_PA)>0, ]

distMat <- vegdist(samples_PA, method = "jaccard")
res <- pcoa(distMat, correction = "none")
res$vectors <- as.data.frame(res$vectors)
res$vectors$Sample_type <- samples$samples[rownames(res$vectors), "Sample_type"]
res$vectors$Country <- as.character(samples$samples[rownames(res$vectors), "Country"])
res$vectors$Flow_regimen <- as.character(samples$samples[rownames(res$vectors), "Flow_regimen"])
```

#### All - Country
```{r, pcoa-by-sample-all-Country, fig.height=10, fig.width=14, warning=F, results='asis', message=F}
g1 <- ggplot(res$vectors,
             aes(x = Axis.1, y = Axis.2, color = Country)) +
  geom_point(size = 2.5) + 
  guides(color=guide_legend(ncol=1)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PCoa (presence/absence) - Bact02")

g2 <- ggplot(res$vectors,
       aes(x = Axis.1, y = Axis.2, color = Country)) +
  geom_point(size = 2.5) + 
  guides(color=guide_legend(ncol=1)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PCoa (presence/absence) - Bact02")

grid.arrange(g1, g2, nrow=2)

rm(g1)
rm(g2)
```

#### All - Sample type
```{r, pcoa-by-sample-all-soil, fig.height=10, fig.width=14, warning=F, results='asis', message=F}
g1 <- ggplot(res$vectors,
             aes(x = Axis.1, y = Axis.2, color = Sample_type)) +
  geom_point(size = 2.5) + 
  guides(color=guide_legend(ncol=1)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PCoa (presence/absence) - Bact02")

g2 <- ggplot(res$vectors,
       aes(x = Axis.1, y = Axis.2, color = Sample_type)) +
  geom_point(size = 2.5) + 
  guides(color=guide_legend(ncol=1)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PCoa (presence/absence) - Bact02")

grid.arrange(g1, g2, nrow=2)

rm(g1)
rm(g2)
```


#### All - Flow regimen
```{r, pcoa-by-sample-all-flow, fig.height=10, fig.width=14, warning=F, results='asis', message=F}
g1 <- ggplot(res$vectors,
             aes(x = Axis.1, y = Axis.2, color = Flow_regimen)) +
  geom_point(size = 2.5) + 
  guides(color=guide_legend(ncol=1)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PCoa (presence/absence) - Bact02")

g2 <- ggplot(res$vectors,
       aes(x = Axis.1, y = Axis.2, color = Flow_regimen)) +
  geom_point(size = 2.5) + 
  guides(color=guide_legend(ncol=1)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PCoa (presence/absence) - Bact02")

grid.arrange(g1, g2, nrow=2)

rm(res)
rm(distMat)
rm(g1)
rm(g2)
```

#### Country {.tabset}
```{r, compare-sample-Presence-absence-Country, fig.height=8, fig.width=10, results='asis', warning=F, message=F}
for(country in sort(unique(samples$samples$Country))){
  cat(paste0("##### ", country, "\n"))
  
  sub_mat <- subset_metabarlist(samples, "samples", samples$samples$Country==country)
  samples_PA <- sub_mat$reads
  samples_PA[samples_PA>0] <- 1
  samples_PA <- samples_PA[, colSums(samples_PA)>1]
  samples_PA <- samples_PA[rowSums(samples_PA)>0, ]
  
  distMat <- vegdist(samples_PA, method = "jaccard")
  res <- pcoa(distMat, correction = "none")
  res$vectors <- as.data.frame(res$vectors)
  res$vectors$Sample_type <- sub_mat$samples[rownames(res$vectors), "Sample_type"]
  res$vectors$Flow_regimen <- as.character(sub_mat$samples[rownames(res$vectors), "Flow_regimen"])
  
  g1 <- ggplot(res$vectors,
         aes(x = Axis.1, y = Axis.2, color = Flow_regimen)) +
    geom_point(size = 2.5) + 
    guides(color=guide_legend(ncol=1)) +
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle(paste0("PCoa (presence/absence) - Bact02 - ", country))
  
  g2 <- ggplot(res$vectors,
         aes(x = Axis.1, y = Axis.2, color = Sample_type)) +
    geom_point(size = 2.5) + 
    guides(color=guide_legend(ncol=1)) +
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle(paste0("PCoa (presence/absence) - Bact02 - ", country))
  
  grid.arrange(g1, g2, nrow=2)
  cat("\n\n")
  
  rm(samples_PA)
  rm(res)
  rm(distMat)
  rm(g1)
  rm(g2)
  rm(sub_mat)
}
```

#### Sample type {.tabset}
```{r, compare-sample-Presence-absence-sample-type, fig.height=8, fig.width=10, results='asis', warning=F, message=F}
for(sample_type in sort(unique(samples$samples$Sample_type))){
  cat(paste0("##### Sample type - ", sample_type, "\n"))
  
  sub_mat <- subset_metabarlist(samples, "samples", samples$samples$Sample_type==sample_type)
  samples_PA <- sub_mat$reads
  samples_PA[samples_PA>0] <- 1
  samples_PA <- samples_PA[, colSums(samples_PA)>1]
  samples_PA <- samples_PA[rowSums(samples_PA)>0, ]
  
  distMat <- vegdist(samples_PA, method = "jaccard")
  res <- pcoa(distMat, correction = "none")
  res$vectors <- as.data.frame(res$vectors)
  res$vectors$Flow_regimen <- sub_mat$samples[rownames(res$vectors), "Flow_regimen"]
  res$vectors$Country <- sub_mat$samples[rownames(res$vectors), "Country"]
  
  g1 <- ggplot(res$vectors,
         aes(x = Axis.1, y = Axis.2, color = Country)) +
    geom_point(size = 2.5) + 
    guides(color=guide_legend(ncol=1)) +
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle(paste0("PCoa (presence/absence) - Bact02 - Sample type: ", sample_type))
  
  g2 <- ggplot(res$vectors,
         aes(x = Axis.1, y = Axis.2, color = Flow_regimen)) +
    geom_point(size = 2.5) + 
    guides(color=guide_legend(ncol=1)) +
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle(paste0("PCoa (presence/absence) - Bact02 - Sample type: ", sample_type))
  
  grid.arrange(g1, g2, nrow=2)
  cat("\n\n")
  
  rm(samples_PA)
  rm(res)
  rm(distMat)
  rm(g1)
  rm(g2)
  rm(sub_mat)
}
```

#### Flow regimen {.tabset}
```{r, compare-sample-Presence-absence-flow, fig.height=8, fig.width=10, results='asis', warning=F, message=F}
for(flow in sort(unique(samples$samples$Flow_regimen))){
  cat(paste0("##### Flow regimen - ", flow, "\n"))
  
  sub_mat <- subset_metabarlist(samples, "samples", samples$samples$Flow_regimen==flow)
  samples_PA <- sub_mat$reads
  samples_PA[samples_PA>0] <- 1
  samples_PA <- samples_PA[, colSums(samples_PA)>1]
  samples_PA <- samples_PA[rowSums(samples_PA)>0, ]
  
  distMat <- vegdist(samples_PA, method = "jaccard")
  res <- pcoa(distMat, correction = "none")
  res$vectors <- as.data.frame(res$vectors)
  res$vectors$Sample_type <- sub_mat$samples[rownames(res$vectors), "Sample_type"]
  res$vectors$Country <- sub_mat$samples[rownames(res$vectors), "Country"]
  
  g1 <- ggplot(res$vectors,
         aes(x = Axis.1, y = Axis.2, color = Country)) +
    geom_point(size = 2.5) + 
    guides(color=guide_legend(ncol=1)) +
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle(paste0("PCoa (presence/absence) - Bact02 - Flow regimen: ", flow))
  
  g2 <- ggplot(res$vectors,
         aes(x = Axis.1, y = Axis.2, color = Sample_type)) +
    geom_point(size = 2.5) + 
    guides(color=guide_legend(ncol=1)) +
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle(paste0("PCoa (presence/absence) - Bact02 - Flow regimen: ", flow))
  
  grid.arrange(g1, g2, nrow=2)
  cat("\n\n")
  
  rm(samples_PA)
  rm(res)
  rm(distMat)
  rm(g1)
  rm(g2)
  rm(sub_mat)
}
```


### classification {.tabset}
#### Country {.tabset}
```{r, hclust-Country, fig.height=20, fig.width=20, results='asis', warning=F, message=F}
for(country in sort(unique(samples$samples$Country))){
  cat(paste0("##### ", country, "\n"))
  
  sub_mat <- subset_metabarlist(samples, "samples", samples$samples$Country==country)
  
  d <- vegdist(decostand(sub_mat$reads, method = "hellinger"), method = "euclidean")
  # Give low weights to variables with low counts and many zeros
  d[is.na(d)] <- 1
  hc1 <- hclust(d, method = "ward.D" )
  tr <- ape::as.phylo(hc1)
  toto <- as.numeric(as.factor(sub_mat$samples[tr$tip.label, "Sample_type"]))
  plot(tr, tip.color = rainbow(6)[toto], cex=0.4, main=paste0("Country ", country), 
       type="fan")
  legend("topleft",
         legend = sort(unique(sub_mat$samples[tr$tip.label, "Sample_type"])), 
         fill = rainbow(6), cex=0.8, text.width=0.485,
         inset=.03, ncol=1,
         box.lty=0)
  
  cat("\n\n")
  
  rm(tr)
  rm(toto)
  rm(hc1)
  rm(sub_mat)
  rm(d)
}
```

#### Sample type {.tabset}
```{r, hclust-sample-type, fig.height=20, fig.width=20, results='asis', warning=F, message=F}
for(sample_type in sort(unique(samples$samples$Sample_type))){
  cat(paste0("##### Sample type - ", sample_type, "\n"))
  
  sub_mat <- subset_metabarlist(samples, "samples", samples$samples$Sample_type==sample_type)
  
  d <- vegdist(decostand(sub_mat$reads, method = "hellinger"), method = "euclidean")
  # Give low weights to variables with low counts and many zeros
  d[is.na(d)] <- 1
  hc1 <- hclust(d, method = "ward.D")
  tr <- ape::as.phylo(hc1)
  toto <- as.numeric(as.factor(sub_mat$samples[tr$tip.label, "Flow_regimen"]))
  plot(tr, tip.color = rainbow(6)[toto], cex=0.4, main=paste0("Sample type: ", sample_type),
       type="fan")
  legend("topleft",
         legend = sort(unique(sub_mat$samples[tr$tip.label, "Flow_regimen"])), 
         fill = rainbow(6), cex=0.8, text.width=0.485,
         inset=.03, ncol=1,
         box.lty=0)
  
  cat("\n\n")
  
  rm(tr)
  rm(toto)
  rm(hc1)
  rm(sub_mat)
  rm(d)
}
```

#### Flow regimen {.tabset}
```{r, hclust-flow, fig.height=20, fig.width=20, results='asis', warning=F, message=F}
for(flow in sort(unique(samples$samples$Flow_regimen))){
  cat(paste0("##### Flow regimen - ", flow, "\n"))
  
  sub_mat <- subset_metabarlist(samples, "samples", samples$samples$Flow_regimen==flow)
  
  d <- vegdist(decostand(sub_mat$reads, method = "hellinger"), method = "euclidean")
  # Give low weights to variables with low counts and many zeros
  d[is.na(d)] <- 1
  hc1 <- hclust(d, method = "ward.D")
  tr <- ape::as.phylo(hc1)
  toto <- as.numeric(as.factor(sub_mat$samples[tr$tip.label, "Sample_type"]))
  plot(tr, tip.color = rainbow(6)[toto], cex=0.4, main=paste0("Flow regimen: ", flow),
       type="fan")
  legend("topleft",
         legend = sort(unique(sub_mat$samples[tr$tip.label, "Sample_type"])), 
         fill = rainbow(6), cex=0.8, text.width=0.485,
         inset=.03, ncol=1,
         box.lty=0)
  
  cat("\n\n")
  
  rm(tr)
  rm(toto)
  rm(hc1)
  rm(sub_mat)
  rm(d)
}
```

### Phylum distribution {.tabset}
```{r, format-data-phylum, eval=T, warning=FALSE, message=FALSE}
bact02_mtbl <- subset_metabarlist(samples, "motus", !samples$motus$bad_taxo_assignment)
bact02_mtbl <- subset_metabarlist(bact02_mtbl, "pcrs" , rowSums(bact02_mtbl$reads)>0)

paths <- format_ecotag_path(bact02_mtbl$motus$path,
                            c("superkingdom", "kingdom", "phylum", "class", 
                              "order", "family", "genus", "species"))

colnames(paths) <- paste0(colnames(paths), "_ecotag")
bact02_mtbl$motus[, colnames(paths)] <- paths
rm(paths)
bact02_mtbl$motus[!is.na(bact02_mtbl$motus$species_name), "species_ecotag"] <- bact02_mtbl$motus[!is.na(bact02_mtbl$motus$species_name), "species_name"]

bact02_mtbl$motus[is.na(bact02_mtbl$motus$phylum_ecotag), "phylum_ecotag"] <- "Unknown_phylum"

phylum_bact02 <- aggregate_motus(bact02_mtbl, bact02_mtbl$motus$phylum_silva)
phylum_bact02$reads <- decostand(phylum_bact02$reads, method = "total", MARGIN = 1) * 100

rm(bact02_mtbl)
rm(mtbl)
rm(samples)
rm(samples_freq)
```

#### Boxplot {.tabset}
##### Sample type {.tabset}
###### Group by phylum
```{r, phylum-plot-group-by-phylum, warning=F, message=F, fig.width=12, fig.height=24}
to_plot <- melt(phylum_bact02$reads)
colnames(to_plot) <- c("Site", "Phylum", "value")
to_plot[, "Sample_type"] <- phylum_bact02$samples[to_plot$Site, "Sample_type"]
to_plot[, "superkingdom"] <- phylum_bact02$motus[to_plot$Phylum, "superkingdom_silva"]

to_plot$Phylum <- factor(to_plot$Phylum, 
                         levels = rownames(rowsum(to_plot[, "value"], 
                                                  to_plot$Phylum))[order(rank(rowsum(to_plot[, "value"], 
                                                                                     to_plot$Phylum)))])

to_plot$superkingdom <- factor(to_plot$superkingdom, 
                          levels = rownames(rowsum(to_plot[, "value"], 
                                                   to_plot$superkingdom))[order(rank(-rowsum(to_plot[, "value"], 
                                                                                        to_plot$superkingdom)))])

ggplot(to_plot, aes(x=value, y=Phylum)) + 
  geom_boxplot(color="grey") + theme_bw() + 
  geom_jitter(alpha=0.4, aes(colour=Sample_type), size=0.8) + 
  scale_x_sqrt(name="Abondance relative (%)") +
  facet_grid(vars(superkingdom), scales = "free_y") + 
  theme(plot.title = element_text(hjust = 0.5),
        strip.text = element_text(size = 7)) +
  guides(color=guide_legend(ncol=1)) +
  ggtitle(paste0("Phylum distribution by sample group by Superkingdom"))

```

###### Group by metadata
```{r, phylum-plot-group-by-soil-type, warning=F, message=F, fig.width=12, fig.height=24}
ggplot(to_plot, aes(x=value, y=Phylum, color=Sample_type)) + 
  geom_boxplot(outlier.size = 0) + theme_bw() + 
  geom_jitter(alpha=0.6, size=0.9, 
              position = position_jitterdodge(jitter.width = 0.15), 
              aes(group = Sample_type)) +
  scale_x_sqrt(name="Abondance relative (%)") +
  facet_grid(vars(superkingdom), scales = "free_y") + 
  theme(plot.title = element_text(hjust = 0.5),
        strip.text = element_text(size = 7)) +
  guides(color=guide_legend(ncol=1)) +
  ggtitle(paste0("Phylum distribution by sample group by sample type and Superkingdom"))
  
rm(to_plot)
```

##### Country {.tabset}
###### Group by phylum
```{r, phylum-plot-group-by-phylum-country, warning=F, message=F, fig.width=12, fig.height=24}
to_plot <- melt(phylum_bact02$reads)
colnames(to_plot) <- c("Site", "Phylum", "value")
to_plot[, "Country"] <- phylum_bact02$samples[to_plot$Site, "Country"]
to_plot[, "superkingdom"] <- phylum_bact02$motus[to_plot$Phylum, "superkingdom_silva"]

to_plot$Phylum <- factor(to_plot$Phylum, 
                         levels = rownames(rowsum(to_plot[, "value"], 
                                                  to_plot$Phylum))[order(rank(rowsum(to_plot[, "value"], 
                                                                                     to_plot$Phylum)))])

to_plot$superkingdom <- factor(to_plot$superkingdom, 
                          levels = rownames(rowsum(to_plot[, "value"], 
                                                   to_plot$superkingdom))[order(rank(-rowsum(to_plot[, "value"], 
                                                                                        to_plot$superkingdom)))])

ggplot(to_plot, aes(x=value, y=Phylum)) + 
  geom_boxplot(color="grey") + theme_bw() + 
  geom_jitter(alpha=0.4, aes(colour=Country), size=0.8) + 
  scale_x_sqrt(name="Abondance relative (%)") +
  facet_grid(vars(superkingdom), scales = "free_y") + 
  theme(plot.title = element_text(hjust = 0.5),
        strip.text = element_text(size = 7)) +
  guides(color=guide_legend(ncol=1)) +
  ggtitle(paste0("Phylum distribution by sample group by Superkingdom"))

```

###### Group by metadata
```{r, phylum-plot-group-by-country_MTD, warning=F, message=F, fig.width=12, fig.height=24}
ggplot(to_plot, aes(x=value, y=Phylum, color=Country)) + 
  geom_boxplot(outlier.size = 0) + theme_bw() + 
  geom_jitter(alpha=0.6, size=0.9, 
              position = position_jitterdodge(jitter.width = 0.15), 
              aes(group = Country)) +
  scale_x_sqrt(name="Abondance relative (%)") +
  facet_grid(vars(superkingdom), scales = "free_y") + 
  theme(plot.title = element_text(hjust = 0.5),
        strip.text = element_text(size = 7)) +
  guides(color=guide_legend(ncol=1)) +
  ggtitle(paste0("Phylum distribution by sample group by country and Superkingdom"))

rm(to_plot)
```


##### Flow regimen {.tabset}
###### Group by phylum
```{r, phylum-plot-group-by-phylum-FR, warning=F, message=F, fig.width=12, fig.height=24}
to_plot <- melt(phylum_bact02$reads)
colnames(to_plot) <- c("Site", "Phylum", "value")
to_plot[, "Flow_regimen"] <- phylum_bact02$samples[to_plot$Site, "Flow_regimen"]
to_plot[, "superkingdom"] <- phylum_bact02$motus[to_plot$Phylum, "superkingdom_silva"]

to_plot$Phylum <- factor(to_plot$Phylum, 
                         levels = rownames(rowsum(to_plot[, "value"], 
                                                  to_plot$Phylum))[order(rank(rowsum(to_plot[, "value"], 
                                                                                     to_plot$Phylum)))])

to_plot$superkingdom <- factor(to_plot$superkingdom, 
                          levels = rownames(rowsum(to_plot[, "value"], 
                                                   to_plot$superkingdom))[order(rank(-rowsum(to_plot[, "value"], 
                                                                                        to_plot$superkingdom)))])

ggplot(to_plot, aes(x=value, y=Phylum)) + 
  geom_boxplot(color="grey") + theme_bw() + 
  geom_jitter(alpha=0.4, aes(colour=Flow_regimen), size=0.8) + 
  scale_x_sqrt(name="Abondance relative (%)") +
  facet_grid(vars(superkingdom), scales = "free_y") + 
  theme(plot.title = element_text(hjust = 0.5),
        strip.text = element_text(size = 7)) +
  guides(color=guide_legend(ncol=1)) +
  ggtitle(paste0("Phylum distribution by sample group by Superkingdom"))

```

###### Group by metadata
```{r, phylum-plot-group-by-FR_MTD, warning=F, message=F, fig.width=12, fig.height=24}
ggplot(to_plot, aes(x=value, y=Phylum, color=Flow_regimen)) + 
  geom_boxplot(outlier.size = 0) + theme_bw() + 
  geom_jitter(alpha=0.6, size=0.9, 
              position = position_jitterdodge(jitter.width = 0.15), 
              aes(group = Flow_regimen)) +
  scale_x_sqrt(name="Abondance relative (%)") +
  facet_grid(vars(superkingdom), scales = "free_y") + 
  theme(plot.title = element_text(hjust = 0.5),
        strip.text = element_text(size = 7)) +
  guides(color=guide_legend(ncol=1)) +
  ggtitle(paste0("Phylum distribution by sample group by flow regimen and Superkingdom"))
  
rm(to_plot)
```

#### Barplot {.tabset}
```{r, barplot-phylum-distribution, echo=F, warning=F, message=F, results='asis', fig.height=8, fig.width=14}
phylum_distribution_function <- function(metabarlist, mtd){
l <- lapply(unique(metabarlist$samples[, mtd]), function(rep) {
  data <- metabarlist$reads[metabarlist$pcrs$sample_id %in% 
                       rownames(metabarlist$samples[metabarlist$samples[, mtd]==rep, ]), ,drop=F]
  as.list(colSums(data) / sum(data))
})

names(l) <- unique(metabarlist$samples[, mtd])
to_plot <- melt(l, value.name="abundance")
names(to_plot) <- c("abundance", "phylum", "metadata")
to_plot$phylum <- factor(to_plot$phylum, 
                         levels = unique(to_plot[order(to_plot$abundance, decreasing = T), "phylum"]))

ggplot(to_plot,
       aes(x=metadata, y=abundance, fill=phylum)) + 
  geom_bar(position="fill", stat="identity", color="black", width=0.7) +
  theme_bw() + 
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, .01))) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle(paste0("Phylum distribution - ", mtd))
}

for(categ in c("Country", "Sample_type", "Flow_regimen")){
  cat(paste0("##### ", categ, "\n"))
  print(phylum_distribution_function(phylum_bact02, categ))
  cat("\n\n")
}
rm(l)
rm(to_plot)
```


```{r}
rm(phylum_bact02)
```


## Clustered motus {.tabset}
To generate all PCA plots, the data are normalized with the Hellinger transformation.

```{r, init-clust-mtbl, warning=F, message=F}
clust_samples <- readRDS(clust_rds_file)
clust_samples <- subset_metabarlist(clust_samples, "pcrs", clust_samples$pcrs$type=="sample")
clust_samples <- aggregate_pcrs(clust_samples)
clust_samples_freq <- clust_samples
clust_samples_freq$reads <- decostand(clust_samples_freq$reads, "hellinger")
```

### Boxplot {.tabset}
#### DNA plate
```{r, boxplot-by-mtd-dnaplate-clust, fig.height=8, fig.width=10, warning=F}
ggBoxPCRMtd(clust_samples, "dna_plate")
```

#### Sample type
```{r, boxplot-by-mtd-soil-clust, fig.height=8, fig.width=8, warning=F}
ggBoxSamplesMtd(clust_samples, "Sample_type")
```

#### Country 
```{r, boxplot-by-mtd-Country-clust, fig.height=8, fig.width=8, warning=F}
ggBoxSamplesMtd(clust_samples, "Country")
```

#### Flow regimen 
```{r, boxplot-by-mtd-flow-clust, fig.height=8, fig.width=8, warning=F}
ggBoxSamplesMtd(clust_samples, "Flow_regimen")
```

### PCA - Abundance {.tabset}
```{r, pca-by-sample-all-clust, fig.height=10, fig.width=14, warning=F, results='asis', message=F}
res.pca <- dudi.pca(clust_samples_freq$reads, scannf = F, nf=4, scale = F)
categories <- clust_samples_freq$samples[rownames(res.pca$li), "Sample_type"]
categories2 <- as.character(clust_samples_freq$samples[rownames(res.pca$li), "Country"])
categories3 <- as.character(clust_samples_freq$samples[rownames(res.pca$li), "Flow_regimen"])
```

#### All - Country
```{r, pca-by-sample-all-Country-clust, fig.height=12, fig.width=14, warning=F, message=F}
g1 <- ggplot(res.pca$li,
       aes(x = Axis1, y = Axis2, color=categories2, shape=categories)) +
  geom_point(size = 2.5) + 
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PCA by sample - Bact02 - Clustered motus") + 
  guides(color=guide_legend(ncol=2))

g2 <- ggplot(res.pca$li,
       aes(x = Axis1, y = Axis3, color=categories2, shape=categories)) +
  geom_point(size = 2.5) + 
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PCA by sample - Bact02 - Clustered motus") + 
  guides(color=guide_legend(ncol=2))

grid.arrange(g1, g2)

rm(g1)
rm(g2)
```

#### All - Sample type
```{r, pca-by-sample-all-soil-clust, fig.height=12, fig.width=14, warning=F, message=F}
g1 <- ggplot(res.pca$li,
       aes(x = Axis1, y = Axis2, color=categories, shape=categories2)) +
  geom_point(size = 2.5) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PCA by sample - Bact02 - Clustered motus") +
  guides(color=guide_legend(ncol=2))


g2 <- ggplot(res.pca$li,
       aes(x = Axis1, y = Axis3, color=categories, shape=categories2)) +
  geom_point(size = 2.5) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PCA by sample - Bact02 - Clustered motus") +
  guides(color=guide_legend(ncol=2))

grid.arrange(g1, g2)

rm(g1)
rm(g2)
```

#### All - Flow regimen
```{r, pca-by-sample-all-flow-regimen-clust, fig.height=12, fig.width=14, warning=F, message=F}
g1 <- ggplot(res.pca$li,
       aes(x = Axis1, y = Axis2, color=categories, shape=categories3)) +
  geom_point(size = 2.5) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PCA by sample - Bact02 - Clustered motus") +
  guides(color=guide_legend(ncol=2))


g2 <- ggplot(res.pca$li,
       aes(x = Axis1, y = Axis3, color=categories, shape=categories2)) +
  geom_point(size = 2.5) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PCA by sample - Bact02 - Clustered motus") +
  guides(color=guide_legend(ncol=2))

grid.arrange(g1, g2)

rm(res.pca)
rm(categories)
rm(categories2)
rm(categories3)
rm(g1)
rm(g2)
```

#### Country {.tabset}
```{r, pca-by-sample-Country-clust, fig.height=12, fig.width=14, warning=F, results='asis', message=F}
for(country in sort(unique(clust_samples_freq$samples$Country))){
  sub_mat <- subset_metabarlist(clust_samples_freq, "samples", clust_samples_freq$samples$Country==country)
  cat(paste0("##### ", country, "\n"))
  res.pca <- dudi.pca(sub_mat$reads, scannf = F, nf=4, scale = F)
  categories <- sub_mat$samples[rownames(res.pca$li), "Flow_regimen"]
  categories2 <- as.character(sub_mat$samples[rownames(res.pca$li), "Sample_type"])
  
  g1 <- ggplot(res.pca$li,
         aes(x = Axis1, y = Axis2, color=categories2, shape=categories)) +
    geom_point(size = 2.5) + 
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle(paste0("PCA by sample - Bact02 - ", country)) + 
    guides(color=guide_legend(ncol=2))
  
  
  g2 <- ggplot(res.pca$li,
         aes(x = Axis1, y = Axis3, color=categories2, shape=categories)) +
    geom_point(size = 2.5) + 
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle(paste0("PCA by sample - Bact02 - ", country)) + 
    guides(color=guide_legend(ncol=2))
  
  grid.arrange(g1, g2)
  cat("\n\n")
  
  rm(res.pca)
  rm(categories)
  rm(categories2)
  rm(g1)
  rm(g2)
  rm(sub_mat)
}
```


#### Sample type {.tabset}
```{r, pca-by-sample-sample-type-clust, fig.height=12, fig.width=14, warning=F, results='asis', message=F}
for(sample_type in sort(unique(clust_samples_freq$samples$Sample_type))){
  sub_mat <- subset_metabarlist(clust_samples_freq, "samples", clust_samples_freq$samples$Sample_type==sample_type)
  cat(paste0("##### Sample type - ", sample_type, "\n"))
  res.pca <- dudi.pca(sub_mat$reads, scannf = F, nf=4, scale = F)
  categories <- sub_mat$samples[rownames(res.pca$li), "Flow_regimen"]
  categories2 <- sub_mat$samples[rownames(res.pca$li), "Country"]
  
  g1 <- ggplot(res.pca$li,
         aes(x = Axis1, y = Axis2, color=categories2, shape=categories)) +
    geom_point(size = 2.5) + 
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle(paste0("PCA by sample - Bact02 - Sample type: ", sample_type)) + 
    guides(color=guide_legend(ncol=2))
  
  
  g2 <- ggplot(res.pca$li,
         aes(x = Axis1, y = Axis3, color=categories2, shape=categories)) +
    geom_point(size = 2.5) + 
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle(paste0("PCA by sample - Bact02 - Sample type: ", sample_type)) + 
    guides(color=guide_legend(ncol=2))
  
  grid.arrange(g1, g2)
  cat("\n\n")
  
  rm(res.pca)
  rm(categories)
  rm(categories2)
  rm(g1)
  rm(g2)
  rm(sub_mat)
}
```

#### Flow regimen {.tabset}
```{r, pca-by-sample-flow-clust, fig.height=12, fig.width=14, warning=F, results='asis', message=F}
for(flow in sort(unique(clust_samples_freq$samples$Flow_regimen))){
  sub_mat <- subset_metabarlist(clust_samples_freq, "samples", clust_samples_freq$samples$Flow_regimen==flow)
  cat(paste0("##### Flow regimen - ", flow, "\n"))
  res.pca <- dudi.pca(sub_mat$reads, scannf = F, nf=4, scale = F)
  categories <- sub_mat$samples[rownames(res.pca$li), "Sample_type"]
  categories2 <- sub_mat$samples[rownames(res.pca$li), "Country"]
  
  g1 <- ggplot(res.pca$li,
         aes(x = Axis1, y = Axis2, color=categories2, shape=categories)) +
    geom_point(size = 2.5) + 
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle(paste0("PCA by sample - Bact02 - Flow regimen: ", flow)) + 
    guides(color=guide_legend(ncol=2))
  
  
  g2 <- ggplot(res.pca$li,
         aes(x = Axis1, y = Axis3, color=categories2, shape=categories)) +
    geom_point(size = 2.5) + 
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle(paste0("PCA by sample - Bact02 - Flow regimen: ", flow)) + 
    guides(color=guide_legend(ncol=2))
  
  grid.arrange(g1, g2)
  cat("\n\n")
  
  rm(res.pca)
  rm(categories)
  rm(categories2)
  rm(g1)
  rm(g2)
  rm(sub_mat)
}
```


### PCoA - Presence-Absence {.tabset}
```{r, pcoa-by-sample-all-clust, fig.height=10, fig.width=14, warning=F, results='asis', message=F}
samples_PA <- clust_samples$reads
samples_PA[samples_PA>0] <- 1
samples_PA <- samples_PA[, colSums(samples_PA)>1]
samples_PA <- samples_PA[rowSums(samples_PA)>0, ]

distMat <- vegdist(samples_PA, method = "jaccard")
res <- pcoa(distMat, correction = "none")
res$vectors <- as.data.frame(res$vectors)
res$vectors$Sample_type <- clust_samples$samples[rownames(res$vectors), "Sample_type"]
res$vectors$Country <- as.character(clust_samples$samples[rownames(res$vectors), "Country"])
res$vectors$Flow_regimen <- as.character(clust_samples$samples[rownames(res$vectors), "Flow_regimen"])
```

#### All - Country
```{r, pcoa-by-sample-all-Country-clust, fig.height=10, fig.width=14, warning=F, results='asis', message=F}
g1 <- ggplot(res$vectors,
             aes(x = Axis.1, y = Axis.2, color = Country)) +
  geom_point(size = 2.5) + 
  guides(color=guide_legend(ncol=1)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PCoa (presence/absence) - Bact02 - clustered motus")

g2 <- ggplot(res$vectors,
       aes(x = Axis.1, y = Axis.2, color = Country)) +
  geom_point(size = 2.5) + 
  guides(color=guide_legend(ncol=1)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PCoa (presence/absence) - Bact02 - clustered motus")

grid.arrange(g1, g2, nrow=2)

rm(g1)
rm(g2)
```

#### All - Sample type
```{r, pcoa-by-sample-all-soil-clust, fig.height=10, fig.width=14, warning=F, results='asis', message=F}
g1 <- ggplot(res$vectors,
             aes(x = Axis.1, y = Axis.2, color = Sample_type)) +
  geom_point(size = 2.5) + 
  guides(color=guide_legend(ncol=1)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PCoa (presence/absence) - Bact02 - clustered motus")

g2 <- ggplot(res$vectors,
       aes(x = Axis.1, y = Axis.2, color = Sample_type)) +
  geom_point(size = 2.5) + 
  guides(color=guide_legend(ncol=1)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PCoa (presence/absence) - Bact02 - clustered motus")

grid.arrange(g1, g2, nrow=2)

rm(g1)
rm(g2)
```


#### All - Flow regimen
```{r, pcoa-by-sample-all-flow-clust, fig.height=10, fig.width=14, warning=F, results='asis', message=F}
g1 <- ggplot(res$vectors,
             aes(x = Axis.1, y = Axis.2, color = Flow_regimen)) +
  geom_point(size = 2.5) + 
  guides(color=guide_legend(ncol=1)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PCoa (presence/absence) - Bact02 - clustered motus")

g2 <- ggplot(res$vectors,
       aes(x = Axis.1, y = Axis.2, color = Flow_regimen)) +
  geom_point(size = 2.5) + 
  guides(color=guide_legend(ncol=1)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PCoa (presence/absence) - Bact02 - clustered motus")

grid.arrange(g1, g2, nrow=2)

rm(res)
rm(distMat)
rm(g1)
rm(g2)
```

#### Country {.tabset}
```{r, compare-sample-Presence-absence-Country-clust, fig.height=8, fig.width=10, results='asis', warning=F, message=F}
for(country in sort(unique(clust_samples$samples$Country))){
  cat(paste0("##### ", country, "\n"))
  
  sub_mat <- subset_metabarlist(clust_samples, "samples", clust_samples$samples$Country==country)
  samples_PA <- sub_mat$reads
  samples_PA[samples_PA>0] <- 1
  samples_PA <- samples_PA[, colSums(samples_PA)>1]
  samples_PA <- samples_PA[rowSums(samples_PA)>0, ]
  
  distMat <- vegdist(samples_PA, method = "jaccard")
  res <- pcoa(distMat, correction = "none")
  res$vectors <- as.data.frame(res$vectors)
  res$vectors$Sample_type <- sub_mat$samples[rownames(res$vectors), "Sample_type"]
  res$vectors$Flow_regimen <- as.character(sub_mat$samples[rownames(res$vectors), "Flow_regimen"])
  
  g1 <- ggplot(res$vectors,
         aes(x = Axis.1, y = Axis.2, color = Flow_regimen)) +
    geom_point(size = 2.5) + 
    guides(color=guide_legend(ncol=1)) +
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle(paste0("PCoa (presence/absence) - Bact02 - clustered motus - ", country))
  
  g2 <- ggplot(res$vectors,
         aes(x = Axis.1, y = Axis.2, color = Sample_type)) +
    geom_point(size = 2.5) + 
    guides(color=guide_legend(ncol=1)) +
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle(paste0("PCoa (presence/absence) - Bact02 - clustered motus - ", country))
  
  grid.arrange(g1, g2, nrow=2)
  cat("\n\n")
  
  rm(samples_PA)
  rm(res)
  rm(distMat)
  rm(g1)
  rm(g2)
  rm(sub_mat)
}
```

#### Sample type {.tabset}
```{r, compare-sample-Presence-absence-sample-type-clust, fig.height=8, fig.width=10, results='asis', warning=F, message=F}
for(sample_type in sort(unique(clust_samples$samples$Sample_type))){
  cat(paste0("##### Sample type - ", sample_type, "\n"))
  
  sub_mat <- subset_metabarlist(clust_samples, "samples", clust_samples$samples$Sample_type==sample_type)
  samples_PA <- sub_mat$reads
  samples_PA[samples_PA>0] <- 1
  samples_PA <- samples_PA[, colSums(samples_PA)>1]
  samples_PA <- samples_PA[rowSums(samples_PA)>0, ]
  
  distMat <- vegdist(samples_PA, method = "jaccard")
  res <- pcoa(distMat, correction = "none")
  res$vectors <- as.data.frame(res$vectors)
  res$vectors$Flow_regimen <- sub_mat$samples[rownames(res$vectors), "Flow_regimen"]
  res$vectors$Country <- sub_mat$samples[rownames(res$vectors), "Country"]
  
  g1 <- ggplot(res$vectors,
         aes(x = Axis.1, y = Axis.2, color = Country)) +
    geom_point(size = 2.5) + 
    guides(color=guide_legend(ncol=1)) +
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle(paste0("PCoa (presence/absence) - Bact02 - clustered motus - Sample type: ", sample_type))
  
  g2 <- ggplot(res$vectors,
         aes(x = Axis.1, y = Axis.2, color = Flow_regimen)) +
    geom_point(size = 2.5) + 
    guides(color=guide_legend(ncol=1)) +
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle(paste0("PCoa (presence/absence) - Bact02 - clustered motus - Sample type: ", sample_type))
  
  grid.arrange(g1, g2, nrow=2)
  cat("\n\n")
  
  rm(samples_PA)
  rm(res)
  rm(distMat)
  rm(g1)
  rm(g2)
  rm(sub_mat)
}
```

#### Flow regimen {.tabset}
```{r, compare-sample-Presence-absence-flow-clust, fig.height=8, fig.width=10, results='asis', warning=F, message=F}
for(flow in sort(unique(clust_samples$samples$Flow_regimen))){
  cat(paste0("##### Flow regimen - ", flow, "\n"))
  
  sub_mat <- subset_metabarlist(clust_samples, "samples", clust_samples$samples$Flow_regimen==flow)
  samples_PA <- sub_mat$reads
  samples_PA[samples_PA>0] <- 1
  samples_PA <- samples_PA[, colSums(samples_PA)>1]
  samples_PA <- samples_PA[rowSums(samples_PA)>0, ]
  
  distMat <- vegdist(samples_PA, method = "jaccard")
  res <- pcoa(distMat, correction = "none")
  res$vectors <- as.data.frame(res$vectors)
  res$vectors$Sample_type <- sub_mat$samples[rownames(res$vectors), "Sample_type"]
  res$vectors$Country <- sub_mat$samples[rownames(res$vectors), "Country"]
  
  g1 <- ggplot(res$vectors,
         aes(x = Axis.1, y = Axis.2, color = Country)) +
    geom_point(size = 2.5) + 
    guides(color=guide_legend(ncol=1)) +
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle(paste0("PCoa (presence/absence) - Bact02 - clustered motus - Flow regimen: ", flow))
  
  g2 <- ggplot(res$vectors,
         aes(x = Axis.1, y = Axis.2, color = Sample_type)) +
    geom_point(size = 2.5) + 
    guides(color=guide_legend(ncol=1)) +
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle(paste0("PCoa (presence/absence) - Bact02 - clustered motus - Flow regimen: ", flow))
  
  grid.arrange(g1, g2, nrow=2)
  cat("\n\n")
  
  rm(samples_PA)
  rm(res)
  rm(distMat)
  rm(g1)
  rm(g2)
  rm(sub_mat)
}
```


### Phylum distribution {.tabset}
```{r, format-data-phylum-clust, eval=T, warning=FALSE, message=FALSE}
bact02_mtbl <- subset_metabarlist(clust_samples, "motus", 
                                  !clust_samples$motus$bad_taxo_assignment)
bact02_mtbl <- subset_metabarlist(bact02_mtbl, "pcrs" , rowSums(bact02_mtbl$reads)>0)

paths <- format_ecotag_path(bact02_mtbl$motus$path,
                            c("superkingdom", "kingdom", "phylum", "class", 
                              "order", "family", "genus", "species"))

colnames(paths) <- paste0(colnames(paths), "_ecotag")
bact02_mtbl$motus[, colnames(paths)] <- paths
rm(paths)
bact02_mtbl$motus[!is.na(bact02_mtbl$motus$species_name), "species_ecotag"] <- bact02_mtbl$motus[!is.na(bact02_mtbl$motus$species_name), "species_name"]

bact02_mtbl$motus[is.na(bact02_mtbl$motus$phylum_ecotag), "phylum_ecotag"] <- "Unknown_phylum"

phylum_bact02 <- aggregate_motus(bact02_mtbl, bact02_mtbl$motus$phylum_silva)
phylum_bact02$reads <- decostand(phylum_bact02$reads, method = "total", MARGIN = 1) * 100

rm(bact02_mtbl)
rm(mtbl)
rm(clust_samples)
rm(clust_samples_freq)
```

#### Boxplot {.tabset}
##### Sample type {.tabset}
###### Group by phylum
```{r, phylum-plot-group-by-phylum-clust, warning=F, message=F, fig.width=12, fig.height=24}
to_plot <- melt(phylum_bact02$reads)
colnames(to_plot) <- c("Site", "Phylum", "value")
to_plot[, "Sample_type"] <- phylum_bact02$samples[to_plot$Site, "Sample_type"]
to_plot[, "superkingdom"] <- phylum_bact02$motus[to_plot$Phylum, "superkingdom_silva"]

to_plot$Phylum <- factor(to_plot$Phylum, 
                         levels = rownames(rowsum(to_plot[, "value"], 
                                                  to_plot$Phylum))[order(rank(rowsum(to_plot[, "value"], 
                                                                                     to_plot$Phylum)))])

to_plot$superkingdom <- factor(to_plot$superkingdom, 
                          levels = rownames(rowsum(to_plot[, "value"], 
                                                   to_plot$superkingdom))[order(rank(-rowsum(to_plot[, "value"], 
                                                                                        to_plot$superkingdom)))])

ggplot(to_plot, aes(x=value, y=Phylum)) + 
  geom_boxplot(color="grey") + theme_bw() + 
  geom_jitter(alpha=0.4, aes(colour=Sample_type), size=0.8) + 
  scale_x_sqrt(name="Abondance relative (%)") +
  facet_grid(vars(superkingdom), scales = "free_y") + 
  theme(plot.title = element_text(hjust = 0.5),
        strip.text = element_text(size = 7)) +
  guides(color=guide_legend(ncol=1)) +
  ggtitle(paste0("Phylum distribution by sample group by Superkingdom - Clustered motus"))

```

###### Group by metadata
```{r, phylum-plot-group-by-soil-type-clust, warning=F, message=F, fig.width=12, fig.height=24}
ggplot(to_plot, aes(x=value, y=Phylum, color=Sample_type)) + 
  geom_boxplot(outlier.size = 0) + theme_bw() + 
  geom_jitter(alpha=0.6, size=0.9, 
              position = position_jitterdodge(jitter.width = 0.15), 
              aes(group = Sample_type)) +
  scale_x_sqrt(name="Abondance relative (%)") +
  facet_grid(vars(superkingdom), scales = "free_y") + 
  theme(plot.title = element_text(hjust = 0.5),
        strip.text = element_text(size = 7)) +
  guides(color=guide_legend(ncol=1)) +
  ggtitle(paste0("Phylum distribution by sample group by sample type and Superkingdom - Clustered motus"))
  
rm(to_plot)
```

##### Country {.tabset}
###### Group by phylum
```{r, phylum-plot-group-by-phylum-country-clust, warning=F, message=F, fig.width=12, fig.height=24}
to_plot <- melt(phylum_bact02$reads)
colnames(to_plot) <- c("Site", "Phylum", "value")
to_plot[, "Country"] <- phylum_bact02$samples[to_plot$Site, "Country"]
to_plot[, "superkingdom"] <- phylum_bact02$motus[to_plot$Phylum, "superkingdom_silva"]

to_plot$Phylum <- factor(to_plot$Phylum, 
                         levels = rownames(rowsum(to_plot[, "value"], 
                                                  to_plot$Phylum))[order(rank(rowsum(to_plot[, "value"], 
                                                                                     to_plot$Phylum)))])

to_plot$superkingdom <- factor(to_plot$superkingdom, 
                          levels = rownames(rowsum(to_plot[, "value"], 
                                                   to_plot$superkingdom))[order(rank(-rowsum(to_plot[, "value"], 
                                                                                        to_plot$superkingdom)))])

ggplot(to_plot, aes(x=value, y=Phylum)) + 
  geom_boxplot(color="grey") + theme_bw() + 
  geom_jitter(alpha=0.4, aes(colour=Country), size=0.8) + 
  scale_x_sqrt(name="Abondance relative (%)") +
  facet_grid(vars(superkingdom), scales = "free_y") + 
  theme(plot.title = element_text(hjust = 0.5),
        strip.text = element_text(size = 7)) +
  guides(color=guide_legend(ncol=1)) +
  ggtitle(paste0("Phylum distribution by sample group by Superkingdom - Clustered motus"))

```

###### Group by metadata
```{r, phylum-plot-group-by-country_MTD-clust, warning=F, message=F, fig.width=12, fig.height=24}
ggplot(to_plot, aes(x=value, y=Phylum, color=Country)) + 
  geom_boxplot(outlier.size = 0) + theme_bw() + 
  geom_jitter(alpha=0.6, size=0.9, 
              position = position_jitterdodge(jitter.width = 0.15), 
              aes(group = Country)) +
  scale_x_sqrt(name="Abondance relative (%)") +
  facet_grid(vars(superkingdom), scales = "free_y") + 
  theme(plot.title = element_text(hjust = 0.5),
        strip.text = element_text(size = 7)) +
  guides(color=guide_legend(ncol=1)) +
  ggtitle(paste0("Phylum distribution by sample group by country and Superkingdom - Clustered motus"))

rm(to_plot)
```


##### Flow regimen {.tabset}
###### Group by phylum
```{r, phylum-plot-group-by-phylum-FR-clust, warning=F, message=F, fig.width=12, fig.height=24}
to_plot <- melt(phylum_bact02$reads)
colnames(to_plot) <- c("Site", "Phylum", "value")
to_plot[, "Flow_regimen"] <- phylum_bact02$samples[to_plot$Site, "Flow_regimen"]
to_plot[, "superkingdom"] <- phylum_bact02$motus[to_plot$Phylum, "superkingdom_silva"]

to_plot$Phylum <- factor(to_plot$Phylum, 
                         levels = rownames(rowsum(to_plot[, "value"], 
                                                  to_plot$Phylum))[order(rank(rowsum(to_plot[, "value"], 
                                                                                     to_plot$Phylum)))])

to_plot$superkingdom <- factor(to_plot$superkingdom, 
                          levels = rownames(rowsum(to_plot[, "value"], 
                                                   to_plot$superkingdom))[order(rank(-rowsum(to_plot[, "value"], 
                                                                                        to_plot$superkingdom)))])

ggplot(to_plot, aes(x=value, y=Phylum)) + 
  geom_boxplot(color="grey") + theme_bw() + 
  geom_jitter(alpha=0.4, aes(colour=Flow_regimen), size=0.8) + 
  scale_x_sqrt(name="Abondance relative (%)") +
  facet_grid(vars(superkingdom), scales = "free_y") + 
  theme(plot.title = element_text(hjust = 0.5),
        strip.text = element_text(size = 7)) +
  guides(color=guide_legend(ncol=1)) +
  ggtitle(paste0("Phylum distribution by sample group by Superkingdom - Clustered motus"))

```

###### Group by metadata
```{r, phylum-plot-group-by-FR_MTD-clust, warning=F, message=F, fig.width=12, fig.height=24}
ggplot(to_plot, aes(x=value, y=Phylum, color=Flow_regimen)) + 
  geom_boxplot(outlier.size = 0) + theme_bw() + 
  geom_jitter(alpha=0.6, size=0.9, 
              position = position_jitterdodge(jitter.width = 0.15), 
              aes(group = Flow_regimen)) +
  scale_x_sqrt(name="Abondance relative (%)") +
  facet_grid(vars(superkingdom), scales = "free_y") + 
  theme(plot.title = element_text(hjust = 0.5),
        strip.text = element_text(size = 7)) +
  guides(color=guide_legend(ncol=1)) +
  ggtitle(paste0("Phylum distribution by sample group by flow regimen and Superkingdom - Clustered motus"))
  
rm(to_plot)
```

#### Barplot {.tabset}
```{r, barplot-phylum-distribution-clust, echo=F, warning=F, message=F, results='asis', fig.height=8, fig.width=14}
phylum_distribution_function <- function(metabarlist, mtd){
l <- lapply(unique(metabarlist$samples[, mtd]), function(rep) {
  data <- metabarlist$reads[metabarlist$pcrs$sample_id %in% 
                       rownames(metabarlist$samples[metabarlist$samples[, mtd]==rep, ]), ,drop=F]
  as.list(colSums(data) / sum(data))
})

names(l) <- unique(metabarlist$samples[, mtd])
to_plot <- melt(l, value.name="abundance")
names(to_plot) <- c("abundance", "phylum", "metadata")
to_plot$phylum <- factor(to_plot$phylum, 
                         levels = unique(to_plot[order(to_plot$abundance, decreasing = T), "phylum"]))

ggplot(to_plot,
       aes(x=metadata, y=abundance, fill=phylum)) + 
  geom_bar(position="fill", stat="identity", color="black", width=0.7) +
  theme_bw() + 
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, .01))) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle(paste0("Phylum distribution - ", mtd))
}

for(categ in c("Country", "Sample_type", "Flow_regimen")){
  cat(paste0("##### ", categ, "\n"))
  print(phylum_distribution_function(phylum_bact02, categ))
  cat("\n\n")
}
rm(l)
rm(to_plot)
```