---
output: 
  html_fragment:
    keep_md: true
    smart: true
    md_extensions: +autolink_bare_uris+tex_math_single_backslash
params:
    # lib_dir: /home/lionnetc/travail/document/DRYVER/bact02
    lib_dir: /bettik/lionnetc/dryver/bact02_merged/run_clean_report
    rds_path: rds_files/bact02_clean_7.rds
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, echo=F}
knitr::opts_chunk$set(echo = FALSE)
```


```{r init-variable, echo=F, message=FALSE, warning=FALSE}
lib_dir <- params$lib_dir
rds_file <- file.path(lib_dir, params$rds_path)
```


```{r, load-dependencies, echo=F, message=FALSE, warning=FALSE, results='hide'}
.libPaths( c( .libPaths(), "/bettik/lionnetc/my_r_libraries/") )
library("metabaR")
library("ggplot2")
library("cowplot")
library("DT")
library("stringr")
library("vegan")
library("ade4")
library("ape")
```

```{r, function, echo=F, warning=F, message=F}
## Function

## Descriptive stat

most_abund_motus_distribution <- function(metabarlist, plot_title, agg_low_motus=T){
  
  if(missing(plot_title)){
      plot_title <- "Distribution of motus"
  }
  
  ## get the positive control value
  ctl_pos_name <- rownames(metabarlist$reads)
  ## remove motus which are not present in the positive control
  ctl_pos <- t(metabarlist$reads[, colSums(metabarlist$reads)>0, drop=F])
  
  ## rename motus by motus_nb_scientific_name
  rownames(ctl_pos) <- paste("motus", 1:length(rownames(ctl_pos)), metabarlist$motus[, "scientific_name"], sep="_")
  
  ## get the relative abundance
  ctl_pos_rel <- as.matrix(sweep(ctl_pos, STATS = colSums(ctl_pos), FUN = "/", MARGIN = 2))
  
  to_plot <- as.data.frame(do.call(rbind, lapply(colnames(ctl_pos_rel), function(ctlname){ 
    return(data.frame("sample_name"=rep(ctlname, length(rownames(ctl_pos_rel[, ctlname, drop=F]))), 
                      "motus"=rownames(ctl_pos_rel[, ctlname, drop=F]),
                      "value"= as.vector(ctl_pos_rel[, ctlname, drop=F]), 
                      "taxon"=gsub("motus_[0-9]+_", "", rownames(ctl_pos_rel[, ctlname, drop=F])),
                      type="motus_distribution",
                      stringsAsFactors = F))
  })))
  if (isTRUE(agg_low_motus)) {
    filter <- to_plot$value<=0.01
    low_motus <- to_plot[which(to_plot$value<=0.01), ]
    to_plot <- to_plot[which(to_plot$value>0.01), ]
    to_plot <- rbind(to_plot, data.frame(aggregate(low_motus["value"],
                                                   by = list(sample_name=low_motus$sample_name), 
                                                   FUN = sum), 
                                         "motus"="other_motus",
                                         "taxon"="other_taxon_under_0.01", 
                                         "type"="motus_distribution"))
  }
  
  read_data <- as.data.frame(do.call(rbind, lapply(colnames(ctl_pos), function(ctlname){ 
    return(data.frame("sample_name"=rep(ctlname, length(rownames(ctl_pos[,ctlname, drop=F]))), 
                      "motus"=rownames(ctl_pos[,ctlname, drop=F]),
                      "value"=as.vector(ctl_pos[,ctlname, drop=F]), 
                      "taxon"=gsub("motus_[0-9]+_", "", rownames(ctl_pos[,ctlname, drop=F])),
                      type="replicate_abundance",
                      stringsAsFactors = F))
  })))
  
  if (isTRUE(agg_low_motus)) {
    low_motus <- read_data[filter, ]
    read_data <- read_data[which(!filter), ]
    read_data <- rbind(read_data, data.frame(aggregate(low_motus["value"], 
                                                       by = list(sample_name=low_motus$sample_name), 
                                                       FUN = sum), 
                                             "motus"="other_motus",
                                             "taxon"="other_taxon_under_0.01", 
                                             "type"="replicate_abundance"))
  }
  
  to_plot <- rbind(read_data, to_plot)
  ## define the control name order to keep order in x-axis
  to_plot$sample_name <- factor(to_plot$sample_name, levels = unique(to_plot$sample_name[order(to_plot$sample_name, decreasing = F)]))
  to_plot$type <- factor(to_plot$type, levels = c("replicate_abundance", 
                                                  "motus_distribution"))
  
  ## ordering the motus value to sort motus in stack barchart for experiment data
  to_plot$motus <- reorder(to_plot$motus, to_plot$value)
  to_plot$motus <- factor(to_plot$motus, levels=rev(levels(to_plot$motus)))
  
  ggplot(to_plot[to_plot$value>0,], aes(x=sample_name, y=value, order=motus, fill=taxon)) +
    geom_bar(inherit.aes = T, stat = "identity", colour="black") +
    facet_wrap("type", scales = "free_y", nrow = 2) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
          plot.title = element_text(hjust = 0.5)) +
    # legend.position = "none") +
    ggtitle(plot_title)
}

identify_disfunctional_PCR <- function(metabarlist, sample_categories=NA, 
                                       threshold=NA, motus_thresh=NA, title="Identify dysfunctionnal pcr", filtering=TRUE){
    to_plot <- data.frame(reads=rowSums(metabarlist$reads), 
                          motus=rowSums(metabarlist$reads>0),
                          type=metabarlist$pcrs[,"type"],
                          stringsAsFactors = F)
    to_plot[to_plot$type=="control", "type"] <- metabarlist$pcrs[to_plot$type=="control", "control_type"]
    
    if(!is.na(sample_categories)){
        to_plot$categories <- metabarlist$samples[metabarlist$pcrs$sample_id, 
                                              sample_categories]
        to_plot[is.na(to_plot$categories),"categories"] <- "no_information"
    }
    to_plot <- to_plot[to_plot$reads>0,]
    g1 <- ggplot(to_plot, aes(x=reads,y=motus, 
                              color=type, 
                              ifelse(is.na(sample_categories), NULL, shape=categories))) + 
    geom_point() + scale_x_log10(name="# reads (log10)") + scale_y_log10(name="# motus (log10)") +
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle(title)
    
    if(!is.na(threshold)){
        g1 <- g1 + geom_vline(xintercept = threshold, color = "red", size=0.5)
    }
    if(!is.na(motus_thresh)){
        g1 <- g1 + geom_hline(yintercept = motus_thresh, color = "red", size=0.5)
    }
    print(g1)
    if(filtering){
        return(rownames(to_plot[to_plot$reads<threshold | to_plot$motus<motus_thresh,]))
    }
}
```

```{r, load-data, warning=FALSE, echo=F, message=F}
mtbl <- readRDS(rds_file)
```


## Remove bad replicate after filtering step {.tabset}
### Filtering by read and motus count
Sample replicates with the same characteristics as the extraction and PCR negative controls are removed.

```{r, remove-low-pcr, warning=F, message=F}

identify_disfunctional_PCR(mtbl, title="Replicate abundance", filtering = FALSE, threshold = 210, motus_thresh = 16)
# remove low abundant replicates
lowAbundRep <- names(rowSums(mtbl$reads))[rowSums(mtbl$reads) < 210 | rowSums(mtbl$reads>0) < 16]
mtbl <- subset_metabarlist(mtbl, "reads", rowSums(mtbl$reads) >= 210 & rowSums(mtbl$reads>0) >= 16)
```


### Filtering by pcr replicate dissimilarity {.tabset}
After the previous filtering steps, some pcr replicates have lost too much motus and no longer look like the other sample replicates. This step checks if pcr always replicates for each sample. To do this the same method uses in the section " PCR reproducibility" is ran on filtered data to check if the previous filtering steps have not remove too much motus and replicates.

```{r, remove-bad-replicate-after-filtering, warning=F, message=F, results='asis', fig.height=20, fig.width=20}
bad_replicate <- unlist(lapply(sort(unique(mtbl$samples$Country)), function(country) {
  cat(paste0("#### ", country, " {.tabset}\n"))
  to_remove <- unlist(lapply(unique(mtbl$samples$Sample_type), function(soil_type){
    cat(paste0("##### ", soil_type, "\n"))
    sub_mat <- subset_metabarlist(mtbl, "samples", mtbl$samples$Country==country & mtbl$samples$Sample_type==soil_type)
    to_remove <- rownames(sub_mat$pcrs[sub_mat$pcrs$sample_id %in% 
                                         names(table(sub_mat$pcrs$sample_id))[table(sub_mat$pcrs$sample_id)<2],])
    sub_mat <- subset_metabarlist(sub_mat, "pcrs", 
                                  !rownames(sub_mat$pcrs) %in% to_remove)
    check_pcr <- metabaR:::pcr_outlier(sub_mat, graphics = FALSE)
    to_remove <- c(to_remove, rownames(check_pcr)[!check_pcr$replicating])
    d <- vegdist(decostand(sub_mat$reads, method = "total"), method = "bray")
    d[is.na(d)] <- 1
    tr <- ape::nj(d)
    col_tips <- as.numeric(as.factor(!check_pcr[tr$tip.label, "replicating"]))
    plot(tr, tip.color = c("black", "red")[col_tips], cex = 0.5, 
         main = paste0("NJ tree of replicates distance for ", soil_type, " of ", country),
         type = "fan")
    edgelabels(text = sprintf("%.2f", tr$edge.length), cex=0.4, frame = "none", adj = c(0,0))
    cat("\n\n")
    rm(sub_mat)
    rm(d)
    gc_stat <- gc()
    return(to_remove)
  }))
  cat("\n\n")
  return(to_remove)
}))

sample_tab <- table(mtbl$pcrs[mtbl$pcrs$type=="sample", "sample_id"])
bad_replicate <- c(bad_replicate, rownames(mtbl$pcrs)[mtbl$pcrs$sample_id %in% 
                                                        names(sample_tab)[sample_tab<2]])
mtbl <- subset_metabarlist(mtbl, "pcrs", !rownames(mtbl$pcrs) %in% bad_replicate)
mtbl <- subset_metabarlist(mtbl, "motus", colSums(mtbl$reads>0) > 1)

mtbl$motus[, "not_clustered"] <- grepl("nnn", mtbl$motus$sequence)

all(!rownames(mtbl$pcrs[mtbl$pcrs$type=="sample",]) %in% bad_replicate)
```

```{r, save-mtbl-wo-disfunctional-pcrs, warning=F, echo=F, eval=T}
saveRDS(mtbl, file = file.path(lib_dir, "rds_files", "dryver_bact02_wo_disfunctional-pcrs.rds"))
```


# Filtering verification
## Positive control composition
```{r, check-positive-control-after-filtering, warning=F, message=F, fig.height=8, fig.width=18}
positive_controls <- subset_metabarlist(mtbl, "pcrs", 
                                        mtbl$pcrs$type=="control" & 
                                          mtbl$pcrs$control_type=="positive")

most_abund_motus_distribution(positive_controls, 
                              "Distribution of motus in positive control replicates", agg_low_motus=FALSE) +
theme(legend.position = "none")

rm(positive_controls)
gc_stat <- gc()
```


## Check replication and contamination {.tabset}
### Tree {.tabset}
```{r, nj-tree-pcr-type, warning=F, message=F, results='asis', fig.height=20, fig.width=20}
for(dna in sort(unique(mtbl$pcrs$dna_plate))){
  cat("#### dna plate ", dna, "\n")
  
  sub_mat <- subset_metabarlist(mtbl, "pcrs", mtbl$pcrs$dna_plate==dna)
  sub_mat <- subset_metabarlist(sub_mat, "reads", rowSums(sub_mat$reads)>0)
  
  d <- vegdist(decostand(sub_mat$reads, method = "total"), method = "bray")
  d[is.na(d)] <- 1
  tr<-ape::nj(d)
  
  col_tips <- sub_mat$pcrs[tr$tip.label, "control_type"]
  col_tips[is.na(col_tips)] <- "sample"
  col_tips <- as.numeric(factor(col_tips, levels = c("pcr", "extraction", "sample", "positive", "sequencing")))
  
  plot(tr, tip.color = rainbow(5)[col_tips], cex = 0.5, 
       main = paste0("NJ tree of replicates distance for dna plate ", dna),
       type = "fan")
  
  cat("\n\n")
  rm(sub_mat)
  rm(tr)
  rm(d)
  rm(col_tips)
  gc_stat <- gc()
}
```

### BCA {.tabset}
To make the Between Class Analysis plot, the data are normalized with the Hellinger method.

```{r, BCA-plot, warning=F, message=F, eval=T, results='asis', fig.height=10, fig.width=11}
for(lib in sort(unique(mtbl$pcrs$library))){
  cat("#### library ", lib, "\n")
  
  sub_mat <- subset_metabarlist(mtbl, "pcrs", mtbl$pcrs$library==lib)
  sub_mat <- subset_metabarlist(sub_mat, "reads", rowSums(sub_mat$reads)>0)
  sub_mat$reads <- decostand(sub_mat$reads, method = "hellinger")
  
  sub_mat_pca <- dudi.pca(sub_mat$reads, scale = FALSE, scannf = FALSE, nf = 100)
  
  col_fac <- sub_mat$pcrs$control_type
  col_fac[is.na(col_fac)] <- sub_mat$samples[sub_mat$pcrs[is.na(col_fac), "sample_id"], "Sample_type"]
  col_fac[col_fac %in% c("extraction", "pcr")] <- "ctl_neg"
  
  sub_mat_bca <- bca(sub_mat_pca, 
                 fac = factor(col_fac), 
                 scannf = FALSE, nf = 2)
  
  print(ggplot(data.frame(sub_mat_bca$ls, sub_mat$pcrs), 
         aes(x=CS1, y=CS2, shape=type, 
             color=factor(col_fac, levels = c("biofilm", "sediment", "positive", "ctl_neg")))) + 
          labs(color="Class") +
    geom_point() + stat_ellipse() +
    ggtitle(paste0("BCA - library ", lib)))
  
  cat("\n\n")
}
```


```{r, remove-gc-pcr-conta, fig.height=8, fig.width=8, warning=FALSE}
rm(gc_stat)
```