---
output: 
  html_fragment:
    keep_md: true
    smart: true
    md_extensions: +autolink_bare_uris+tex_math_single_backslash
  #md_document:
    #variant: markdown+autolink_bare_uris+tex_math_single_backslash

params:
    lib_dir: /bettik/lionnetc/dryver/bact02_merged/run_clean_report
    rds_path: rds_file_name
    out_file: NULL
---

```{r setup, include=FALSE, echo=F}
knitr::opts_chunk$set(echo = TRUE)
```


```{r init-variable, echo=F}
lib_dir <- params$lib_dir
rds_file <- file.path(lib_dir, params$rds_path)
if(is.null(params$out_file)){
	out_file <- file.path(lib_dir, "check_pcr")
}else{
	out_file <- file.path(lib_dir, params$out_file)
}
```


```{r, load-dependencies, echo=F, results='hide', message=F, warning=F}
.libPaths( c( .libPaths(), "/bettik/lionnetc/my_r_libraries/") )
library("metabaR")
library("ggplot2")
library("cowplot")
library("ade4")
library("vegan")
```


```{r, load-data, warning=FALSE, echo=F}
mtbl <- readRDS(rds_file)
```


```{r, set-titles, warning=FALSE, echo=F, results='asis'}
if(as.numeric(unique(mtbl$samples$Campaign)) > 0){
	section_title <- paste0("Campaign ", as.character(unique(mtbl$samples$Campaign)))
	subsection_title <- as.character(unique(mtbl$samples$Sample_type))
	cat(paste0("\n### ", section_title, " {.tabset}\n\n"))
	cat(paste0("\n#### ", subsection_title, "\n\n"))
}else{
	cat("\n### Positive Controls - Soil \n\n")
}
```

##### Density plot {.tabset}
```{r, pcrslayer, warning=FALSE, eval=T}
check_pcr <- metabaR:::pcr_outlier(mtbl, graphics = TRUE)
saveRDS(check_pcr, out_file)
```

##### {-}

**Explanation:**

* This plot shows the density distribution of two distances (within and between samples) :
  + the blue line corresponds to the between distance density
  + the black line corresponds to the within distance density

* The red line is the threshold define by the intersection of the two density. The algorithm iterates until the threshold is stabilized.


##### PCoA plot
```{r, pcr-replicate-plot, fig.height=12, fig.width=8, warning=FALSE, eval=T}
a <- check_pcr_repl(mtbl, funcpcr = check_pcr$replicating,
               groups = mtbl$samples[mtbl$pcrs$sample_id, "Country"]) + 
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle(paste0("PCoA before pcr filtering - campaign ", 
		 as.character(unique(mtbl$samples$Campaign))," - ",
		 as.character(unique(mtbl$samples$Sample_type))))

mtbl <- subset_metabarlist(mtbl, "pcrs", check_pcr$replicating)

b <- check_pcr_repl(mtbl, 
               groups = mtbl$samples[mtbl$pcrs$sample_id, "Country"]) + 
  theme(plot.title = element_text(hjust = 0.5), legend.position = "none") +
  ggtitle(paste0("PCoA after pcr filtering - campaign",
		 as.character(unique(mtbl$samples$Campaign))," - ",
                 as.character(unique(mtbl$samples$Sample_type))))

plot_grid(a, b, nrow = 2)

rm(check_pcr)
rm(a)
rm(b)
```
