---
output: 
  html_fragment:
    keep_md: true
    smart: true
    md_extensions: +autolink_bare_uris+tex_math_single_backslash

params:
    lib_dir: /bettik/lionnetc/dryver/euka02_merged/run_clean_report
    rds_path: rds_files/euka02_wo_no_rep_pcr_5.rds
---

```{r setup, include=FALSE, echo=F}
knitr::opts_chunk$set(echo = FALSE)
```


```{r init-variable, echo=F, message=FALSE, warning=FALSE}
lib_dir <- params$lib_dir
rds_file <- file.path(lib_dir, params$rds_path)

rev_tag <- "acacacac,acagcaca,gtgtacat,tatgtcag,tagtcgca,tactatac,actagatc,gatcgcga,cgctctcg,gtcgtaga,gtcacgtc,gactgatg,agactatg,gcgtcagc,tgacatca,acatgtgt,gtacgact,atgatcgc,acgacgag,catcagtc,atcagtca,tctactga,gatgatct,ctgcgtac,agcgacta,tcagtgtc,actctgct,atatagcg,ctatgcta,tcgcgctg,agcacagt,tagctagt,agtgctac,cgtataca,cgagtcgt,cacatgat"

fwd_tag <- "acacacac,acagcaca,gtgtacat,tatgtcag,tagtcgca,tactatac,actagatc,gatcgcga,cgctctcg,gtcgtaga,gtcacgtc,gactgatg,agactatg,gcgtcagc,tgacatca,acatgtgt,gtacgact,atgatcgc,acgacgag,catcagtc,atcagtca,tctactga,gatgatct,ctgcgtac,agcgacta,tcagtgtc,actctgct,atatagcg,ctatgcta,tcgcgctg,agcacagt,tagctagt"

tag_list <- union(unlist(strsplit(fwd_tag, ",")), 
                  unlist(strsplit(rev_tag, ",")))
```


```{r, load-dependencies, echo=F, message=FALSE, warning=FALSE, results='hide'}
.libPaths( c( .libPaths(), "/bettik/lionnetc/my_r_libraries/") )
library("metabaR")
library("ggplot2")
library("cowplot")
library("ggrepel")
library("gridExtra")
library("ade4")
library("RColorBrewer")
library("DT")
library("stringr")
library("vegan")
library("ape")
library("reshape2")
```


```{r, load-data, warning=FALSE, echo=F, message=FALSE}
mtbl <- readRDS(rds_file)
```

Potential contaminants are identified with the enrichment test of Kolmogorov-Smirnov. The test compares the composition of PCR and extraction negative control replicates to the sample replicates. The test is considered significant (alpha risk set to 0.05) if a MOTU is more abundant in PCR negative controls than in samples replicates.

```{r, identify-conta, fig.height=8, fig.width=8, warning=FALSE}
# mtbl <- contaslayer(mtbl, control_types=c("pcr", "extraction"), 
#                       method="all", output_col="not_a_contaminant")

kolmo_test <- function(i) {
  ctl_neg <- !is.na(mtbl$pcrs$control_type) & mtbl$pcrs$control_type %in% c('pcr', 'extraction')
  ks.test(x=mtbl$reads[!ctl_neg, i], 
          y=mtbl$reads[ctl_neg, i], 
          alternative = 'greater')$p.value
  }

## Apply bonferoni correction
nb_motus <- nrow(mtbl$motus)
mtbl$motus$Kolmo.contaminant.p.value <- unlist(lapply(1:nb_motus, kolmo_test)) * nb_motus
mtbl$motus$Kolmo.contaminant <- mtbl$motus$Kolmo.contaminant.p.value < 0.05
rm(nb_motus)

ctl <- subset_metabarlist(mtbl, "pcrs", 
                          mtbl$pcrs$type=="control" &
                            mtbl$pcrs$control_type %in% c('pcr', 'extraction'))

agg_ctl <- aggregate_pcrs(ctl, replicates = ctl$pcrs$control_type)

mtbl$motus[,"in_extraction_control"] <- rownames(mtbl$motus) %in%
  names(agg_ctl$reads["extraction", agg_ctl$reads["extraction",]>0]) 

mtbl$motus[,"in_pcr_control"] <- rownames(mtbl$motus) %in% 
  names(agg_ctl$reads["pcr", agg_ctl$reads["pcr",]>0])

ctl_abund <- t(apply(mtbl$reads, 2, 
                     function(col){ 
                       col[!is.na(mtbl$pcrs$control_type) & 
                             mtbl$pcrs$control_type %in% c("pcr", "extraction")]}))

rownames(ctl_abund) <- colnames(mtbl$reads)

colnames(ctl_abund) <- paste("control_abundance", 
                             colnames(ctl_abund), 
                             sep = ":")

mtbl$motus <- cbind(mtbl$motus,
                    ctl_abund[rownames(mtbl$motus),
                              sort(colnames(ctl_abund))])

rm(ctl)
rm(ctl_abund)
rm(agg_ctl)
```

### PCR template of pcr contaminant {.tabset}
```{r, pcr-template-conta, fig.height=8, fig.width=8, warning=F, echo=FALSE, results ="asis"}
conta <- subset_metabarlist(mtbl, "motus", 
                            mtbl$motus$Kolmo.contaminant==TRUE)

saveRDS(object = conta, file = file.path(lib_dir, "rds_files", "euka02_pcr-ext_conta_6.rds"))

for(lib in unique(conta$pcrs$library)){
  mtb_lib <- subset_metabarlist(conta, "pcrs", conta$pcrs$library==lib)
  cat(paste0("#### ",lib,"\n"))
    
  print(ggpcrtag(mtb_lib, taglist = as.character(tag_list), legend_title = "Reads count") + 
          ggtitle(paste0("PCR template of PCR and extraction contaminants for library ", lib, " - Euka02")) + 
          theme(plot.title = element_text(hjust = 0.5)))
  cat("\n\n")
  rm(mtb_lib)
  gc_stat <- gc()
}
rm(gc_stat)
rm(conta)
```

### Table of PCR controls composition
```{r, table-of-conta, warning=FALSE, results='asis'}
ctl <- subset_metabarlist(mtbl, "pcrs", 
                          mtbl$pcrs$type=="control" & 
                            mtbl$pcrs$control_type %in% c("pcr", "extraction"))

ctl_info <- ctl$motus[names(colSums(ctl$reads)[colSums(ctl$reads)>0]),]

out <- data.frame("scientific_name"=ctl_info$scientific_name,
                  "rank"= ctl_info$rank,
                  "best_identity"=ctl_info$`best_identity:Euka02.DB`,
                  "is_contaminant"=ctl_info$Kolmo.contaminant==TRUE,
                  "in extraction control"=ctl_info$in_extraction_control,
                  "in pcr control"=ctl_info$in_pcr_control,
                  "# reads in all control"=colSums(ctl$reads)[rownames(ctl_info)],
                  "# reads in pcr control"=colSums(ctl$reads[ctl$pcrs$control_type=="pcr",])[rownames(ctl_info)],
                  "# reads in extraction control"=colSums(ctl$reads[ctl$pcrs$control_type=="extraction",])[rownames(ctl_info)],
                  "# reads total"=colSums(mtbl$reads)[rownames(ctl_info)],
                  "% whole dataset"=scales::percent(colSums(mtbl$reads)[rownames(ctl_info)]/sum(colSums(mtbl$reads)), 
                                                    accuracy = 0.01),
                  "sequence"=ctl_info$sequence,
                  check.names = F)


DT::datatable(out[order(out$`# reads in all control`, decreasing = T),],
              extensions = 'Buttons', 
              options = list(dom = 'Bfrtip', 
                             buttons = c('excel', "csv"),
                             scrollX='400px'))

rm(ctl_info)
rm(ctl)
rm(out)
gc_stat <- gc()
```


### PCR plate without pcr and extraction contaminants {.tabset}
```{r, remove-contaminant, fig.height=8, fig.width=8, warning=FALSE}
mtbl <- subset_metabarlist(mtbl, "motus",
                              mtbl$motus$Kolmo.contaminant==FALSE)
```

```{r, save-mtbl-after-conta, fig.height=8, fig.width=8, warning=FALSE}
saveRDS(object = mtbl, file = file.path(lib_dir, "rds_files", "euka02_wo_pcr-ext_conta_6.rds"))
# mtbl <- readRDS(file.path(lib_dir, "rds_files", "euka02_wo_pcr-ext_conta_6.rds"))
```

```{r, pcr-template-without-conta, fig.height=8, fig.width=8, warning=FALSE, results='asis'}
for(lib in unique(mtbl$pcrs$library)){
    mtb_lib <- subset_metabarlist(mtbl, "pcrs", mtbl$pcrs$library==lib)
    
    cat(paste0("#### ",lib,"\n"))
    print(ggpcrtag(mtb_lib, taglist = as.character(tag_list), legend_title = "Reads count") + 
            ggtitle(paste0("PCR template without pcr and extraction contaminants for  library ", lib, " - Euka02")) + 
            theme(plot.title = element_text(hjust = 0.5)))
    cat("\n\n")
    
    rm(mtb_lib)
    gc_stat <- gc()
}
```

```{r, remove-gc-conta, fig.height=8, fig.width=8, warning=FALSE}
rm(gc_stat)
```